\chapter*{Preface}

This notebook is on \textit{Calculus}, a very important mathematical tool that was invented back in Newton's time or even earlier. It has now become an entrance level module for mathematics and engineering background students in their fist year in the university.

Initially, the invention of calculus, including the introduction of differentiation and integration, is of course used to explain things such as the concept of ``speed'' as a differential of distance over time. You might easily come up with common use cases of calculus, for example calculating the tangent of a curve, and calculating the volume of an arbitrarily shaped container. Other applications which may not make too much sense for beginners, for example the derivation of cycloid, are also done using calculus. Many advanced mathematical tools themselves are built on top of calculus, for example fourier transform, which is widely used in signal processing. Without a solid understanding of calculus, it is hardly possible for one to use these tools confidently and effectively.

The key references of this notebook is listed below. During the development of the notebook, this list may become longer and longer.

Book \textit{Calculus Metric Version Eighth Edition} by James Stewart, published by Cengage Learning \cite{stewart2015calculus}.

Book \textit{Calculus} by Gibert Strang (Massachusetts Institute of Technology), published by Wellesley-Cambridge Press \cite{strangt1991calculus}. This book is available at MIT Open Courseware (\textit{ocw.mit.edu}). There are countless number of great learning materials there. 