\chapter*{Preface}

Artificial intelligence (AI) and specifically machine learning were originally a research topic under the scope of control systems, and it is primarily used for system identification (``machine learning was used to be called system identification'', as some professors say). Its artificial neuron network (ANN) structure is a promising approach for building highly nonlinear self-tuning functions.

Due to the limitations of computational power and data storage in the past years, training large-size ANN with massive data points was hardly possible. With the advancement of computer and material science in the beginning of the 21st centry, in particular the development of GPU and TPU, nowadays we can manage networks with dozens of layers, each containing hundreds of neurons. This is referred as the deep learning (DL) network structure. With DL as the building block, effective convolutional neural networks (CNN) and recurrent neural networks (RNN) have been proposed which are effective at finding trends in spacial and sequential data. The proposition of transformers, yet another attention-based deep learning structure, has taken natural language processing to the next level.

AI has grown so significantly in the past decades that it is now considered a separate area apart from classic control systems, and it is drawing more and more attentions. To put it in perspective, while a well-known textbook on control systems may have 10,000+ citations records after years since its publication, a famous conference paper on modern AI can easily have 50,000+ citations after a few months.

This notebook does not focus on the introduction of AI mechanisms, as most of which are already covered in control systems related notebooks. The purpose of this notebook is mainly to keep up-to-date with the latest technology in machine learning, starting from transformer and its applications in chatbot, the most famous of which being ChatGPT developed by OpenAI. The notebook cares not only the techniques, but also the business models used in these technologies and companies.

Some part of this notebook is written by ChatGPT-4.