\chapter*{Preface}

This notebook introduces probability, statistics, data science and engineering. They are the ``must-have'' ability in most, if not all, academic and industrial projects.

In Part I of the notebook, probability theory is introduced. Probability theory studies how likely an event is to occur. It offers rich models and tools to describe random values and stochastic events.

In Part II of the notebook, statistics is introduced. Statistics is a collection of methods to analyze and observe insights from data, verify statistics hypothesis and draw conclusions and predictions.

In Part III of the notebook, commonly used software and toolkits for statistics analysis and data science are introduced. Different from Parts I and II which focus more on theory, Part III focuses more on the tools to solve practical problems. Notice that Artificial Intelligence (AI) has become increasingly popular in recent years for data analysis. There is a separate notebook on introducing AI. In this notebook, the tools to process data using AI is only briefly introduced, again focusing only on the usage of software, not theory.

As a bonus, in Part IV, semantic web, the database framework defined under Web 3.0, is introduced. Semantic web does not necessarily contribute to solving a specific problem using a specific statistics theory. It is rather a concept or organizing and exchanging data efficiently.

Key references of this notebook are summarized as follows. Notice that these materials are so very widely used here and there in the entire notebook that it becomes improbable to address them each time they are used.

For probability and statistics parts:
\begin{itemize}
  \item Spiegel, Murray, John Schiller, and Alu Srinivasan. \textit{Probability and statistics.} 2020.
  \item Dekking, Frederik Michel, et al., \textit{A Modern Introduction to Probability and Statistics: Understanding why and how.} Vol. 488. London: Springer, 2005.
\end{itemize}
For data science part:
\begin{itemize}
  \item Kirill Eremenko, \textit{R Programming A-Z: R For Data Science With Real Exercises}, Udemy Course.
  \item Lakshmanan, Valliappa, Martin Görner, and Ryan Gillard. Practical Machine Learning for Computer Vision. " O'Reilly Media, Inc.", 2021.
  \item Jose Portilla, \textit{Complete TensorFlow 2 and Keras Deep Learning Bootcamp}, Udemy
\end{itemize}

Online materials such as tutorials from YouTube, Bilibili, etc., are also used in forming this notebook. ChatGPT-4 is used as a consultant in forming this notebook. 